Adversarial Training of Neural Networks\\ against Systematic Uncertainty

### 2016-08-29 tex/paper.tex
added algorithm

@@

Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
Let consider the multivariate random variables $X_\lambda: \Omega \mapsto
\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_\lambda$ depends on a
nuisance parameter $\lambda \in \Lambda$ whose values define a
parameterized family of its systematic uncertainties. That is, $X_\lambda$ and
$Y$ induce together a joint probability distribution $p(X,Y|\lambda)$, where the
conditional on $\lambda$ denotes $X_\lambda$. For training, let further assume a
finite set $\{ x_i, y_i, \lambda_i \}_{i=1}^N$ of realizations
$X_{\lambda_i}(\omega_i), Y(\omega_i)$, for $\omega_i \in \Omega$ and known
values $\lambda_i$ of the nuisance parameter. Our goal is to learn a function
$f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal Y}$ of parameters $\theta_f$
(e.g., a neural network-based classifier if ${\cal Y}$ is a finite set of
classes) and minimizing  a loss ${\cal L}_f(\theta_f)$. In addition, we require
that $f(X_\lambda ; \theta_f)$ should be robust to the value of the nuisance parameter $\lambda$ --
which remains unknown at test time. More specifically, we aim at building $f$
such that in the ideal case
\begin{equation}\label{eqn:criterion-true}
f(X_{\lambda_i}(\omega) ; \theta_f) = f(X_{\lambda_j}(\omega) ; \theta_f)
\end{equation} for any
sample $\omega \in \Omega$ and any $\lambda_i, \lambda_j$ pair of values of the
nuisance parameter.

Since we do not have training tuples $(X_{\lambda_i}(\omega),
X_{\lambda_j}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{equation}\label{eqn:criterion-measure}
    P(\{ \omega | f(X_{\lambda_i}(\omega) ; \theta_f) = y \} ) = P( \{ \omega' | f(X_{\lambda_j}(\omega') ; \theta_f) = y\}) \text{ for all $y \in {\cal Y}$}.
\end{equation}
In words, we are looking for a predictive function $f$ which is a pivotal
quantity \citep{degroot1986probability} with respect to the nuisance parameter.
That is, such that  the distribution of $f(X_\lambda; \theta_f)$ is invariant
with respect to $\lambda$. Note that a function $f$ for which
Eqn.~\ref{eqn:criterion-true} is true necessarily satisfies
Eqn.~\ref{eqn:criterion-measure}. The converse is however in general not true, since the
sets of samples $\{ \omega | f(X_{\lambda_i}(\omega); \theta_f) = y \}$ and $\{
\omega' | f(X_{\lambda_j}(\omega'); \theta_f) = y \}$ do not need to be the same
for the equality to hold.
With this distinction, but as only Eqn.~\ref{eqn:criterion-measure} is
of direct interest in this work, we further simplify notations by denoting
the pivotal quantity criterion as
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) | \lambda_i ) = p(f(X ; \theta_f) | \lambda_j ) \text{ for all $\lambda_i,\lambda_j \in  \Lambda$}
\end{equation}.

@@

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary model $r(\cdot ;
\theta_r) : \mathbb{R} \mapsto \Lambda$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_r)$. This model takes as input realizations of $f(X; \theta_f)$
and produces as output estimates $r(f(X; \theta_f)) = \hat{p}(\lambda | f(X; \theta_f)) $
that $f(X; \theta_f)$ is generated from the nuisance value $\lambda$. If $p(f(X; \theta_f)|\lambda)$ varies with $\lambda$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)|\lambda)$ is invariant with $\lambda$ as we require, then $r$
should perform poorly. Training $f$ such that it additionally minimizes the performance of $r$
therefore acts as a regularization towards Eqn.~\ref{eqn:criterion}.

As for generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which we carry out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_r)
\end{equation}
that we optimize by finding the saddle point $(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ such that
\begin{align}
    \smash{\hat\theta_f} &= \arg \min_{\theta_f} E(\theta_f, \smash{\hat\theta_r}), \\
    \smash{\hat\theta_r} &= \arg \max_{\theta_r} E(\smash{\hat\theta_f}, \theta_r).
\end{align}

\glnote{Sketch the optimization algoritm}

@@

>>>>

\usepackage{algorithm}
\usepackage{algpseudocode}

@@

Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
Let consider the multivariate random variables $X_\lambda: \Omega \mapsto
\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_\lambda$ depends on a
nuisance parameter $\lambda \in \Lambda$ whose values define a
parameterized family of its systematic uncertainties. That is, $X_\lambda$ and
$Y$ induce together a joint probability distribution $p(X,Y|\lambda)$, where the
conditional on $\lambda$ denotes $X_\lambda$. For training, let further assume a
finite set $\{ x_i, y_i, \lambda_i \}_{i=1}^N$ of realizations
$X_{\lambda_i}(\omega_i), Y(\omega_i)$, for $\omega_i \in \Omega$ and known
values $\lambda_i$ of the nuisance parameter. Our goal is to learn a function
$f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal Y}$ of parameters $\theta_f$
(e.g., a neural network-based classifier if ${\cal Y}$ is a finite set of
classes) and minimizing  a loss ${\cal L}_f(\theta_f)$. In addition, we require
that $f(X_\lambda ; \theta_f)$ should be robust to the value of the nuisance parameter $\lambda$ --
which remains unknown at test time. More specifically, we aim at building $f$
such that in the ideal case
\begin{equation}\label{eqn:criterion-true}
f(X_{\lambda}(\omega) ; \theta_f) = f(X_{\lambda^\prime}(\omega) ; \theta_f)
\end{equation} for all
samples $\omega \in \Omega$ and all $\lambda, \lambda^\prime$ pairs of values of the
nuisance parameter.

Since we do not have training tuples $(X_{\lambda}(\omega),
X_{\lambda^\prime}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{equation}\label{eqn:criterion-measure}
    P(\{ \omega | f(X_{\lambda}(\omega) ; \theta_f) = y \} ) = P( \{ \omega' | f(X_{\lambda^\prime}(\omega') ; \theta_f) = y\}) \text{ for all $y \in {\cal Y}$}.
\end{equation}
In words, we are looking for a predictive function $f$ which is a pivotal
quantity \citep{degroot1986probability} with respect to the nuisance parameter.
That is, such that  the distribution of $f(X_\lambda; \theta_f)$ is invariant
with respect to the value of $\lambda$. Note that a function $f$ for which
Eqn.~\ref{eqn:criterion-true} is true necessarily satisfies
Eqn.~\ref{eqn:criterion-measure}. The converse is however in general not true, since the
sets of samples $\{ \omega | f(X_{\lambda}(\omega); \theta_f) = y \}$ and $\{
\omega' | f(X_{\lambda^\prime}(\omega'); \theta_f) = y \}$ do not need to be the same
for the equality to hold.
In order to simplify notations,
and as only Eqn.~\ref{eqn:criterion-measure} is
of direct interest in this work, we denote from here on
the pivotal quantity criterion as
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) | \lambda ) = p(f(X ; \theta_f) | \lambda^\prime ) \text{ for all $\lambda,\lambda^\prime \in  \Lambda$.}
\end{equation}

@@

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary classifier $r(\cdot ;
\theta_r) : \mathbb{R} \mapsto \Lambda$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_r)$.
Assuming that $\Lambda$ defines a finite family of nuisance values $\lambda_l$ (for $l=1, \dots, |\Lambda|$),
this model takes as input realizations of $f(X; \theta_f)$
and produces as output probability estimates $r(f(X; \theta_f); \theta_r)_l = \hat{p}(\lambda_l| f(X; \theta_f)) $
that $f(X; \theta_f)$ is generated from the nuisance value $\lambda_l$. If $p(f(X; \theta_f)|\lambda)$ varies with $\lambda$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)|\lambda)$ is invariant with $\lambda$, as we require, then $r$
should perform poorly and be close to random guessing. Training $f$ such that it additionally minimizes the performance of $r$
therefore acts as a regularization towards Eqn.~\ref{eqn:criterion}.

As for generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which we carry out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_r)
\end{equation}
that we optimize by finding the saddle point $(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ such that
\begin{align}
    \smash{\hat\theta_f} &= \arg \min_{\theta_f} E(\theta_f, \smash{\hat\theta_r}), \label{eqn:min_thetaf} \\
    \smash{\hat\theta_r} &= \arg \max_{\theta_r} E(\smash{\hat\theta_f}, \theta_r) \label{eqn:max_thetar}.
\end{align}
The adversarial training procedure to obtain $(\smash{\hat\theta_f},
\smash{\hat\theta_r})$ is formally presented in
Algorithm~\ref{alg:adversarial-training} in the case of $f$ being a classifier
and of using the log-loss for both ${\cal L}_f$ and ${\cal L}_r$. The algorithm
consists in using stochastic gradient descent alternatively to optimize
Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar}.

\begin{algorithm}[t]
\caption{Adversarial training of a classifier $f$ against an adversary $r$.\\
{\it Inputs:} training data $\{ x_i, y_i, \lambda_i \}_{i=1}^N$\\
{\it Outputs:} $\smash{\hat\theta_f}, \smash{\hat\theta_r}$\\
{\it Hyper-parameters:} Number $T$ of training iterations, Number $K$ of gradient steps to update $r$.}
\label{alg:adversarial-training}
\begin{algorithmic}[1]
    \For{$t=1$ to $T$}
        \For{$k=1$ to $K$} \Comment{Update $r$}
            \State{Sample minibatch $\{x_m, \lambda_m \}_{m=1}^M$} of size $M$;
            \State{With $\theta_f$ fixed, update the adversary $r$ by ascending its stochastic gradient
            $$\nabla_{\theta_r} \sum_{m=1}^M \left[ -\sum_{l=1}^{|\Lambda|} 1(\lambda_m = \lambda_l)\log r(f(x_m;\theta_f);\theta_r)_l \right];$$}
        \EndFor
        \State{Sample minibatch $\{x_m, y_m, \lambda_m \}_{m=1}^M$} of size $M$; \Comment{Update $f$}
        \State{With $\theta_r$ fixed, update $f$ by descending its stochastic gradient
        $$\nabla_{\theta_f} \sum_{m=1}^M \left[ \sum_{c=1}^{|{\cal Y}|} 1(y_m = y_c) \log f(x_m;\theta_f)_c -\sum_{l=1}^{|\Lambda|} 1(\lambda_m = \lambda_l)\log r(f(x_m;\theta_f);\theta_r)_l \right];$$}
    \EndFor
\end{algorithmic}
\end{algorithm}

@@

In this section, we show that Algorithm~\ref{alg:adversarial-training} converges
to a classifier $f$ which is a pivotal quantity in the sense of Eqn.~\ref{eqn:criterion}.
